---
title: "Transcriptional Dynamics of Milk Produiction in Bovine Mammary Glands"
author: "Benji Lamp"
date: "4/24/24"
output: 
  html_document: 
    toc: true
    toc_float: true
    theme: united
    highlights: tango
---

# Introduction
It is understood that the mammary gland in cattle undergoes extensive morphological and physiological transformations to facilitate the production of milk. These changes are regulated at the genetic and epigenetic level, with specific patterns of expression driving maturation of the gland. This study aims to uncover the temporal expression patterns of mRNA in bovine mammary glands by performing timewise differential expression of sequencing data, then clustering based on temporal expression pattern.

# Setup {.tabset}

### Libraries
```{r, message=FALSE, warning=FALSE}
library(DESeq2)
library(dplyr)
library(AnnotationDbi)
library(org.Bt.eg.db)
library(sva)
library(readxl)
library(tidyr)
library(knitr)
library(kableExtra)
```

### Data
Import raw count files from github
```{r}
file_path <- 'https://raw.githubusercontent.com/blamp23/Cattle-lac-proj/main/Raw_data/Raw_Cattle_Counts.tabular'
df <- read.delim(file_path, header = TRUE, row.names = 1)
df <- df[-1]
```


# Filtering/loading {.tabset}

## Metadata
Creation of a dataframe that holds information about the experiment, as well as the batch effects
```{r}
samples <- data.frame(
  row.names = c("v_468", "v_502", "v_504", "v_507", "v_509", "v_598", "v_610", 
                "mp_468", "mp_502", "mp_504", "mp_507", "mp_509", "mp_598", "mp_610",
                "lp_468", "lp_502", "lp_504", "lp_509", "lp_507", "lp_598", "lp_610",
                "el_468", "el_502", "el_504", "el_507", "el_509", "el_598", "el_610",
                "l_468", "l_502", "l_504", "l_507", "l_509", "l_598", "l_610"),
  condition = c(rep("v", 7), rep("mp", 7), rep("lp", 7), rep("el", 7), rep("l", 7)),
  animal = c(468, 502, 504, 507, 509, 598, 610,
             468, 502, 504, 507, 509, 598, 610,
             468, 502, 504, 509, 507, 598, 610,
             468, 502, 504, 507, 509, 598, 610,
             468, 502, 504, 507, 509, 598, 610),
  batch = c(1, 1, 1, 1, 1, 1, 1, 
            0, 1, 0, 1, 0, 0, 0,
            0, 1, 0, 0, 1, 0, 0,
            0, 1, 0, 1, 0, 0, 0, 
            1, 1, 1, 1, 1, 1, 1 ),
  BCM = c(1, 1, 1, 1, 1, 1, 1,
          0, 0, 0, 0, 0, 0, 0,
          0, 0, 0, 0, 0, 0, 0,
          0, 0, 0, 0, 0, 0, 0,
          1, 1, 1, 1, 1, 1, 1)
)

new_colnames <- c(row.names(samples))
colnames(df) <- new_colnames

colnames(df)[28] <- "lp_610"  # lp_610 and el_610 are swapped samples
colnames(df)[21] <- "el_610"
samples <- samples[colnames(df), ]
tnames <- c(rep(1, 7), rep(2, 7), rep(3, 7), rep(4, 7), rep(5, 7))
samples$time <- tnames
```

## Filtering
```{r}
# Filter Method A ##############################################################
# originaly 27607 genes
# filtered to 17344 genes
# Step 1: Splitting Data by Condition
conditions <- unique(samples$condition)
split_data <- lapply(conditions, function(cond) {
  df[, samples$condition == cond]
})

# Step 2: Define Filtering Function
count_high_expressions <- function(data) {
  apply(data, 1, function(gene) {
    sum(gene >= 3) >= 4  # Returns TRUE if at least 4 out of 7 samples have counts ≥ 3
  })
}
high_exp_counts <- lapply(split_data, count_high_expressions)

# Step 3: Aggregate Results for inclusion
gene_inclusion_filter <- rowSums(do.call(cbind, high_exp_counts)) >= 1 # If 1/5 timepoints meet crit, gene is included

# Step 4: Filter Genes
df <- df[gene_inclusion_filter, ]
```

## Filtered Counts
```{r, echo=FALSE, results='asis'}
library(knitr)
library(kableExtra)

# Custom CSS to enlarge table captions
cat("
<style>
table caption {
  font-size: 16pt;
  font-weight: bold;
}
</style>
")

# Assuming 'samples' is another data frame you want to display
kable(samples, caption = "Sample Data") %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"), full_width = FALSE)

# Displaying the first few rows of df
df_head <- head(df)

# Splitting df into two halves if it has many columns
half_col <- ceiling(ncol(df_head) / 2)
df_head_first_part <- df_head[, 1:half_col]
df_head_second_part <- df_head[, (half_col+1):ncol(df_head)]

# Display the first half
kable(df_head_first_part, caption = "Filtered Counts") %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"), full_width = FALSE)

# Display the second half
kable(df_head_second_part) %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"), full_width = FALSE)
```













