---
title: "Transcriptional Dynamics of Milk Production in Bovine Mammary Glands"
author: "Benji Lamp"
date: "4/24/24"
output: 
  html_document: 
    toc: true
    toc_float: true
    theme: united
    highlights: tango
---

# Introduction

It is understood that the mammary gland in cattle undergoes extensive morphological and physiological transformations to facilitate the production of milk. These changes are regulated at the genetic and epigenetic level, with specific patterns of expression driving maturation of the gland. This study aims to uncover the temporal expression patterns of mRNA in bovine mammary glands by performing timewise differential expression of sequencing data, then clustering based on temporal expression pattern.

# Galaxy 
All paired end reads were loaded into galaxy as single end. After preliminary QC was performed, alignment to the bosTau9 reference genome was performed. After alignment, statistics were again drawn in order to ensure quality of reads. Mapping to genome to generate counts was performed by Feature counts. Read mapping took in to account the strandedness of the reads, All stranded reads were read in as forward stranded. In reference to the metadata, the column 'batch' delimitates which samples were stranded and unstranded (1,0 respectively). 


# Setup {.tabset}

### Libraries

```{r, message=FALSE, warning=FALSE}
library(DESeq2)
library(dplyr)
library(AnnotationDbi)
library(org.Bt.eg.db)
library(sva)
library(readxl)
library(tidyr)
library(knitr)
library(kableExtra)
library(ggplot2)
library(plotly)
library(ggrepel)

```

### Data

The raw count files can be downloaded directly from Github. The data was loaded using the read.delim function from the base package, setting the Ensembl ids as rownames. The first column, which was redundant (gene-ids), was removed.

```{r}
file_path <- 'https://raw.githubusercontent.com/blamp23/Cattle-lac-proj/main/Raw_data/Raw_Cattle_Counts.tabular'
df <- read.delim(file_path, header = TRUE, row.names = 1)
df <- df[-1]
```

# Filtering/loading

## Metadata

Sample metadata was compiled to reflect the stages of pregnancy as well as the batches associated with strandedness. The metadata table includes the animal identifier, stage of pregnancy, and batch for each sample.

```{r, echo=FALSE}
samples <- data.frame(
  row.names = c("v_468", "v_502", "v_504", "v_507", "v_509", "v_598", "v_610", 
                "mp_468", "mp_502", "mp_504", "mp_507", "mp_509", "mp_598", "mp_610",
                "lp_468", "lp_502", "lp_504", "lp_509", "lp_507", "lp_598", "lp_610",
                "el_468", "el_502", "el_504", "el_507", "el_509", "el_598", "el_610",
                "l_468", "l_502", "l_504", "l_507", "l_509", "l_598", "l_610"),
  condition = c(rep("v", 7), rep("mp", 7), rep("lp", 7), rep("el", 7), rep("l", 7)),
  animal = c(468, 502, 504, 507, 509, 598, 610,
             468, 502, 504, 507, 509, 598, 610,
             468, 502, 504, 509, 507, 598, 610,
             468, 502, 504, 507, 509, 598, 610,
             468, 502, 504, 507, 509, 598, 610),
  batch = c(1, 1, 1, 1, 1, 1, 1, 
            0, 1, 0, 1, 0, 0, 0,
            0, 1, 0, 0, 1, 0, 0,
            0, 1, 0, 1, 0, 0, 0, 
            1, 1, 1, 1, 1, 1, 1 )
)
```

Before finalizing the raw counts, it is necessary to adjust the dataframe's column names to simplify data manipulation. The desired format combines the condition with the sample identifier (e.g., XX_123), which mirrors the row names in the metadata dataframe. The format mimics the row names of the metadata dataframe. Additionally, it is important to note that the samples from Early Lactation 610 and Late Pregnancy 610 were switched prior to sequencing and must be corrected before proceeding with the analysis.

```{r, echo=F, results='asis', message=FALSE}
new_colnames <- c(row.names(samples))
colnames(df) <- new_colnames

colnames(df)[28] <- "lp_610"  # lp_610 and el_610 are swapped samples
colnames(df)[21] <- "el_610"

el_610 <- which(names(df) == "el_610")
lp_610 <- which(names(df) == "lp_610")

# Create a new vector of column names with the positions of these two columns swapped
new_order <- names(df)
new_order[c(el_610, lp_610)] <- new_order[c(lp_610, el_610)]

# Reorder the columns in the DataFrame
df <- df[, new_order]

samples <- samples[colnames(df), ]



# Custom CSS to enlarge table captions
cat("
<style>
table caption {
  font-size: 16pt;
  font-weight: bold;
}
</style>
")

# Assuming 'samples' is another data frame you want to display
kable(samples, caption = "Sample Data") %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"), full_width = FALSE)

```

## Filtering

To determine which genes should be included in the study, each gene is evaluated within each lactation stage based on its expression count across the 7 samples in that stage. A gene is considered for inclusion in a specific lactation stage if its count exceeds a threshold of 3 in at least 4 of the 7 samples at that time point.

A gene will be included in the final study if it meets the above criteria in at least 1 out of the 5 time points (i.e., lactation stages). This means that a gene must have high counts in at least one of the lactation stages to be included in the study.

```{r}
# Step 1: Splitting Data by Condition
conditions <- unique(samples$condition)
split_data <- lapply(conditions, function(cond) {
  df[, samples$condition == cond]
})

# Step 2: Define Filtering Function
count_high_expressions <- function(data) {
  apply(data, 1, function(gene) {
    sum(gene >= 3) >= 4  # Returns TRUE if at least 4 out of 7 samples have counts ≥ 3
  })
}
high_exp_counts <- lapply(split_data, count_high_expressions)

# Step 3: Aggregate Results for inclusion
gene_inclusion_filter <- rowSums(do.call(cbind, high_exp_counts)) >= 1 # If 1/5 timepoints meet crit, gene is included

# Step 4: Filter Genes
df <- df[gene_inclusion_filter, ]
```

# Normalization

## ComBatSeq

Strandedness as a batch effect was seen to have distorted our data when using principal component analysis as a diagnostic tool. In order to correct the batch effect, the ComBat_seq function from the sva package was used. This function adjusts for batch effects using the batch variable from our metadata and considers the experimental design (like time course variables) to prevent overfitting.

```{r, message=FALSE}
count_matrix <- as.matrix(df)

cdf <- ComBat_seq(count_matrix,
                  batch = samples$batch,
                  group = samples$condition)
```

## DESeq2


A standard DESeq function was applied using the counts corrected by ComBat_seq and the experimental design from the 'samples' metadata frame.

```{r, message=FALSE, warning=FALSE}
dds <- DESeqDataSetFromMatrix(
  countData = cdf,
  colData = samples,
  design = ~ condition
)

dds <- DESeq(dds)
```


```{r, message = F, warning=F, echo = F}

dds <- DESeq(dds)
res <- results(dds)


dds1 <- DESeqDataSetFromMatrix(
  countData = count_matrix,
  colData = samples,
  design = ~ condition
)

dds1 <- DESeq(dds1)
```

### PCA {.tabset}

#### Corrected PCA
```{r, message=FALSE, warning=FALSE, echo=F}
vsd <- vst(dds)
# Get the PCA data
pcaData <- plotPCA(vsd, intgroup=c("condition"), returnData=TRUE)

# Extract the percent variance captured by each principal component
percentVar <- round(100 * attr(pcaData, "percentVar"))

# Plot with ggplot2
ggplot(pcaData, aes(x=PC1, y=PC2, color=condition)) +
  geom_point(size=3) +
  geom_text_repel(aes(label=row.names(pcaData))) +  # using ggrepel to prevent overlap
  stat_ellipse(aes(group=condition), level=0.85) +  # adding ellipses around each condition group
  xlab(paste0("PC1: ", percentVar[1], "% variance")) +
  ylab(paste0("PC2: ", percentVar[2], "% variance")) +
  theme_bw() +
  ggtitle("ComBat Corrected PCA Plot")

```

#### Raw PCA
```{r, message=FALSE, warning=FALSE, echo=F}
vsd <- vst(dds1)
# Get the PCA data
pcaData <- plotPCA(vsd, intgroup=c("condition"), returnData=TRUE)

# Extract the percent variance captured by each principal component
percentVar <- round(100 * attr(pcaData, "percentVar"))

# Plot with ggplot2
ggplot(pcaData, aes(x=PC1, y=PC2, color=condition)) +
  geom_point(size=3) +
  geom_text_repel(aes(label=row.names(pcaData))) +  # using ggrepel to prevent overlap
  stat_ellipse(aes(group=condition), level=0.85) +  # adding ellipses around each condition group
  xlab(paste0("PC1: ", percentVar[1], "% variance")) +
  ylab(paste0("PC2: ", percentVar[2], "% variance")) +
  theme_bw() +
  ggtitle("Uncorrected PCA")

```

# Timecourse Analysis

```{r, echo=FALSE, message=FALSE}
dds <- estimateSizeFactors(dds)
normalized_counts_df <- as.data.frame(counts(dds, normalized=TRUE))
col_names <- rep(c("v", "mp", "lp", "el", "l"), each = 7)
names(normalized_counts_df) <- col_names

# Prepare an empty data frame for storing averages, with correct dimensions and column names
averages_df <- data.frame(matrix(ncol = length(unique(col_names)), nrow = nrow(normalized_counts_df)))
names(averages_df) <- unique(col_names)
rownames(averages_df) <- rownames(normalized_counts_df)

# Loop through each time-point to calculate and store row means
for (timepoint in unique(col_names)) {
  # Identify columns belonging to the current time-point
  cols <- which(col_names == timepoint)
  # Calculate and store
  averages_df[[timepoint]] <- rowMeans(normalized_counts_df[, cols], na.rm = TRUE)
}

```

```{r, echo=FALSE, message=FALSE, results='asis'}
hadf <- head(averages_df)
# Custom CSS to enlarge table captions
cat("
<style>
table caption {
  font-size: 16pt;
  font-weight: bold;
}
</style>
")

# Assuming 'samples' is another data frame you want to display
kable(hadf, caption = "Mean of Normalized Counts") %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"), full_width = FALSE)
```

## Pairwise Comparison

To perform differential expression analysis, comparative structures need to be formed between timepoints. The loop below iterates through each successive comparison and generates 'res' objects from DESeq2, which contain statistics like log fold change and adjusted p-values for each comparison. After the False Discovery Rate adjusted p-value (padj) is generated, the mean of the normalized counts is evaluated between each comparison. If gene expression increases from time n to n+1, an 'I' will be assigned to the gene. If gene expression decreases from n to n+1, a 'D' will be assigned to the gene.

The adjusted p-value generated from DESeq2 is then used to evaluate whether there is significance. If the padj value is greater than 0.05, the gene will be deemed to have an insignificant change in expression between timepoints and will be assigned an 'S'



```{r}
contrasts <- list(
  c("condition", "l", "v"),
  #
  c("condition", "el", "v"),
  c("condition", "l", "mp"),
  #
  c("condition", "lp", "v"),
  c("condition", "el", "mp"),
  c("condition", "l", "lp"),
  #
  c("condition", "mp", "v"),
  c("condition", "lp", "mp"),
  c("condition", "el", "lp"),
  c("condition", "l", "el")
)

contrast_results <- list()

# DE for pairs in contrasts
for (i in seq_along(contrasts)) {
  contrast = contrasts[[i]]
  res <- results(dds,
                 contrast=contrast)
  
  # Determine if the mean expression increases or decreases
  condition1_mean <- averages_df[[contrast[3]]]  # Mean for the first condition
  condition2_mean <- averages_df[[contrast[2]]]  # Mean for the second condition
  expressionDirection <- ifelse(condition1_mean < condition2_mean, "I", "D")
  
  # Hypothesis test, FTR null assign 'S', Rej null assign I/D
  res$expressionChange <- ifelse(res$padj > 0.05, "S", expressionDirection)
  
  contrast_results[[paste(contrast[3], "to", contrast[2])]] <- res
}
```

## Motif Structuring
The variables (I,S,D) that corespond to expression are currently being stored in the list of DESeq objects. In order to create an index of gene expression, genes are added to a dataframe that completley describes their patterns of expression over the time course. This tabe can be manipulated in order to generate a singular index of pattern structure depending on the needs of the research. 


The complete time course analysis can be organized into a piecewise hierarchical structure with 4 tiers, 

- Primary:		V-PL\
- Secondary:	V-EL MP-PL\
- Tertiary:		V-LP MP-EL LP-PL\
- Quaternary:	V-MP MP-LP LP-EL EL-PL\

A complete model vector will be structured as\
Primary-Secondary-Tertiary-Quaternary\
X-XX-XXX-XXXX\


```{r}
combined_results <- data.frame(gene=rownames(contrast_results[[1]]))
# Loop through each contrast result to combine them
for (i in seq_along(contrast_results)) {
  contrast_name <- names(contrast_results)[i]  # Get the name of the current contrast
  # Create a temporary dataframe with genes and their expression changes for the current contrast
  temp_df <- data.frame(gene=rownames(contrast_results[[i]]), 
                        expressionChange=contrast_results[[i]]$expressionChange)
  colnames(temp_df)[2] <- contrast_name  # Rename the second column to the current contrast name
  # Merge the temporary dataframe with the combined_results dataframe
  combined_results <- merge(combined_results, temp_df, by="gene", all=TRUE)
}

# Concatenate expression patterns to make model vector
combined_results$modelVector <- apply(combined_results[, -1], 1, 
                                      function(x) paste(x, collapse = ""))

# for Quaternary tree structure ###############################################################
combined_results$primary <- combined_results[,2]
combined_results$secondary <- paste0(combined_results[,3], combined_results[,4])
combined_results$tertiary <- paste0(combined_results[,5], combined_results[,6], combined_results[,7])
combined_results$quaternary <- paste0(combined_results[,8], combined_results[,9], combined_results[,10], combined_results[,11])



```

```{r, echo=FALSE, message=FALSE, results='asis'}
relevant_data <- combined_results %>%
  dplyr::select(starts_with("v to"), starts_with("mp to"), starts_with("lp to"), starts_with("el to"))

# Reshape data from wide to long format
long_data <- relevant_data %>%
  pivot_longer(cols = everything(), names_to = "transition", values_to = "event") %>%
  group_by(transition, event) %>%
  summarise(count = n(), .groups = 'drop')

# Spread the data back to wide format to see counts for each event type in each transition
wide_data <- long_data %>%
  pivot_wider(names_from = event, values_from = count, values_fill = list(count = 0))


wide <- as.data.frame(wide_data)

cat("
<style>
table caption {
  font-size: 8pt;
}
</style>
")

# Assuming 'samples' is another data frame you want to display
kable(wide, caption = "Incedence of Increasing, Decreasing and Steady Genes") %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"), full_width = FALSE)
```

```{r, echo=FALSE, message=FALSE, results='asis'}
selected_columns <- combined_results[, c(1, 12:16)]

# Custom CSS to enlarge table captions
cat("
<style>
table caption {
  font-size: 8pt;
  font-weight: bold;
}
</style>
")

# Assuming 'samples' is another data frame you want to display
kable(head(combined_results[,1:11]), caption = "") %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"), full_width = FALSE)

kable(head(selected_columns), caption = "First 5 Rows of Pattern Dataframe") %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"), full_width = FALSE)


```

## Mapping Specific Motif

Reducing the number of comparisons within a pattern structure can significantly decrease the total number of patterns identified. Additionally, depending on the objectives of the experiment, different patterns can be constructed that reveal more biologically relevant structures. Below, the 'vector' variable was constructed to create patterns from virgin to pregnancy to lactation.

```{r}
motif_index <- list()
x <- combined_results
vector <- paste0(x$`v to mp`,
                 x$`v to lp`,
                 x$`mp to el`,
                 x$`lp to el`,
                 x$`mp to l`,
                 x$`lp to l`,
                 x$`el to l`)

# Loop through each unique modelVector
for(model in unique(vector)) {
  # Subset the genes that match the current modelVector
  genes <- combined_results$gene[vector == model]
  # Store
  motif_index[[model]] <- genes
}

# Specific patterns can be called using this function 
head(motif_index$SIIIIIS)
```

Throughout this entire process, Ensembl IDs have been used as gene identifiers. To make the data easier to read, mapping to symbols can be done using AnnotationDbi and the org.Bt.eg.db package for bovine gene symbols.

```{r, message=F, warning=F}
mapped_motif_index <- list()
for(model in names(motif_index)) {
  ensembl_ids <- motif_index[[model]]
  gene_symbols <- tryCatch({
    mapIds(org.Bt.eg.db,
           keys = ensembl_ids,
           column = "SYMBOL",
           keytype = "ENSEMBL",
           multivals = 'first')
  }, error = function(e) {
    message("Error with model ", model, ": ", e$message)
    return(NA)  # Return NA or any other placeholder that indicates an error
  })
  mapped_motif_index[[model]] <- gene_symbols
}
head(mapped_motif_index$SIIIIIS)

```

# Session Info
```{r}
sessionInfo()
```


